France-wide 10\,m Deciduous–Evergreen Mapping: AlphaEarth Embeddings vs Harmonics

### 2025-09-17 article/article_v3.tex
new pass with deepresearch

\keyFont{\section{Keywords:} deciduous-evergreen, AlphaEarth embeddings, harmonic analysis, Random Forest, phenology}
\end{abstract>

\section{Introduction}
An annual deciduous-evergreen layer at 10\,m is missing from European products yet essential for forest disturbance detection. Copernicus DLT achieves 96.5\% accuracy classifying broadleaf/conifer but cannot distinguish deciduous oak from evergreen oak \citep{EU2024a}; BD Forêt V2 provides accurate species-level mapping but remains static, compiled from 2007-2018 aerial photography \citep{IGN2024}. Accurate phenology baselines are critical for reducing false positives in disturbance detection—deciduous forests' natural high variance versus evergreens' stable signals fundamentally affects detection thresholds. Species-specific vulnerabilities (15 insect species and 10 diseases affect coniferous forests versus only 6 insects and 2 diseases in deciduous) require precise phenological discrimination for early warning systems.

Foundation models in Earth observation have emerged as transformative technologies (2022-2025), with AlphaEarth \citep{AlphaEarth2025} exemplifying this paradigm shift. AlphaEarth's 64-dimensional embedding field model, trained on 512 TPU v4 nodes processing 3+ billion observation frames with multi-objective loss functions, achieves 23.9\% error reduction while providing 16× storage efficiency compared to conventional approaches. Already deployed by the UN Food and Agriculture Organization, Harvard Forest, MapBiomas, and Stanford University for ecosystem mapping, it demonstrates operational readiness across diverse applications. Google processes 1.4 trillion embedding footprints annually, democratizing access through Earth Engine without requiring researchers to manage specialized TPU infrastructure. Yet questions persist about their performance versus domain-specific methods for specialized tasks like phenological classification.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/France Phenology Map RF.png}
    \caption{France's 2023 10\,m deciduous-evergreen forest map reveals phenological patterns invisible in leaf-type classifications. Orange areas indicate deciduous forests, cyan shows evergreen forests. Both harmonic and embedding approaches produce similar spatial patterns, with embeddings providing slightly improved boundary delineation in mixed stands.}
    \label{fig:national_map}
\end{figure}

@@

Both approaches address a critical need for forest disturbance detection systems. Modern disturbance detection algorithms struggle to separate seasonal variations from genuine ecological changes, with false positive rates exceeding 20\%. Current methods like BEAST \citep{Zhao2019} and CCDC \citep{Zhu2014} cannot distinguish between natural phenological variations and true disturbances without accurate phenological baselines. Our annual phenology maps provide the critical baseline information these algorithms require—enabling focus on anomalous changes while accounting for expected seasonal patterns that static forest inventories cannot provide. Knowing baseline phenological regimes (deciduous high-variance versus evergreen stability) enables reduction of false positives by 20-30\% in operational systems.

### Transferability and Scalability Considerations

@@

Future applications leverage these annual phenology maps for enhanced disturbance detection, where knowing baseline phenological regimes (deciduous high-variance versus evergreen stability) reduces false positives by 20-30\%. Integration with existing detection algorithms (CCDC, BFAST) enables separation of genuine disturbances from seasonal patterns. Species-level refinement will support targeted pest and disease monitoring, as different forest types show distinct vulnerability profiles—15 insect species and 10 diseases affecting coniferous forests versus only 6 insects and 2 diseases in deciduous forests. The operational efficiency demonstrated here enables continental-scale monitoring with near real-time disturbance alerts.

@@

>>>>

\keyFont{\section{Keywords:} deciduous-evergreen, AlphaEarth embeddings, harmonic analysis, Random Forest, phenology}
\end{abstract}

\section{Introduction}
An annual deciduous-evergreen layer at 10\,m is missing from European products yet essential for forest disturbance detection. Copernicus DLT achieves 96.5\% accuracy classifying broadleaf/conifer but cannot distinguish deciduous oak from evergreen oak \citep{EU2024a}; BD Forêt V2 provides accurate species-level mapping but remains static, compiled from 2007-2018 aerial photography \citep{IGN2024}. Accurate phenology baselines are critical for reducing false positives in disturbance detection—deciduous forests' natural high variance versus evergreens' stable signals fundamentally affects detection thresholds. Species-specific vulnerabilities (15 insect species and 10 diseases affect coniferous forests versus only 6 insects and 2 diseases in deciduous) require precise phenological discrimination for early warning systems. As a pixel-level seasonal baseline, an annual phenology map complements time-series change detection (e.g., CCDC, BEAST) by constraining alerts during expected leaf-off/leaf-on periods and improving discrimination of genuine disturbances \citep{Zhu2014,Zhao2019}.

Foundation models in Earth observation have emerged as transformative technologies (2022-2025), with AlphaEarth \citep{AlphaEarth2025} exemplifying this paradigm shift. AlphaEarth's 64-dimensional embedding field model, trained on 512 TPU v4 nodes processing 3+ billion observation frames with multi-objective loss functions, achieves 23.9\% error reduction while providing 16× storage efficiency compared to conventional approaches.

The versatility of AlphaEarth embeddings extends across diverse Earth observation domains. In agricultural applications, embeddings achieve within-region crop classification accuracies of 91.4-99.1\% across cool climate regions including Northern Europe, Japan, and the United States \citep{CoolClimateAg2025}, outperforming traditional spectral indices through implicit pattern recognition. Urban and land change detection applications demonstrate balanced accuracies above 78-79\%, surpassing conventional alternatives \citep{AlphaEarthFoundations2024}. The system effectively processes Ecuador's persistent cloud-covered agricultural plots and reveals Canadian agricultural variations invisible to conventional analysis, demonstrating robust performance across challenging environments.

Disaster response and environmental monitoring leverage the pre-computed nature of embeddings for rapid deployment. The annual global dataset (2017-2024) at 10-meter resolution democratizes access to high-performance geospatial analysis, enabling practitioners to build monitoring systems for food security, conservation, and disaster response with unprecedented efficiency \citep{SatelliteEmbedding2025}. Google processes embedding footprints for continental-scale analyses that previously required specialized infrastructure, democratizing access through Earth Engine without requiring researchers to manage TPU infrastructure.

Already deployed by the UN Food and Agriculture Organization, Harvard Forest, MapBiomas, and Stanford University for ecosystem mapping, AlphaEarth demonstrates operational readiness across these diverse applications. Yet questions persist about their performance versus domain-specific methods for specialized tasks like phenological classification, particularly regarding interpretability and cross-regional transferability where some studies show performance degradation in transfer scenarios \citep{CoolClimateAg2025}.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/France_Phenology_Map_RF.png}
    \caption{France's 2023 10\,m deciduous-evergreen forest map reveals phenological patterns invisible in leaf-type classifications. Orange areas indicate deciduous forests, cyan shows evergreen forests. Both harmonic and embedding approaches produce similar spatial patterns, with embeddings providing slightly improved boundary delineation in mixed stands.}
    \label{fig:national_map}
\end{figure}

@@

\subsection{Phenology-Informed Disturbance Detection}
Automated change detection methods can confuse seasonal phenology with true disturbances. Decomposition and harmonic frameworks (e.g., CCDC, BEAST) explicitly model or remove seasonal components before testing for breaks \citep{Zhu2014,Zhao2019}. A phenology map provides complementary, external context that further reduces false positives:

\textit{Dynamic thresholds by phenology class} — In deciduous areas with high natural variability, require larger or more sustained vegetation index drops to trigger alarms; in evergreen areas with low variability, allow more sensitive thresholds. This aligns alert sensitivity with each pixel’s phenological noise floor.

\textit{Phenology-aware filtering} — Down-weight or discard alerts that coincide with expected events (e.g., leaf-off in late autumn for pixels mapped as deciduous), adding a rule-based sanity check on top of time-series residual tests.

\textit{Class-specific model configuration} — Tune detection parameters per phenology class (e.g., number of harmonics, residual thresholds), rather than one-size-fits-all settings, to reflect different seasonal dynamics.

\textit{Context-aware sensor fusion} — In deciduous zones, deprioritize optical greenness signals during winter and rely more on complementary cues (e.g., radar backscatter) when validating alerts; in evergreen zones, treat moderate greenness drops as more informative.

Together, these phenology-informed strategies improve precision by filtering out cyclical dynamics while preserving sensitivity to genuine disturbances—allowing operational systems to focus on anomalies that deviate from both modeled seasonality and the mapped phenological regime.

\subsection{Transferability and Scalability Considerations}

@@

Future applications leverage these annual phenology maps for enhanced disturbance detection, where knowing baseline phenological regimes (deciduous high-variance versus evergreen stability) can materially reduce false positives. Integration with existing detection algorithms (CCDC, BFAST) enables separation of genuine disturbances from seasonal patterns. Species-level refinement will support targeted pest and disease monitoring, as different forest types show distinct vulnerability profiles—15 insect species and 10 diseases affecting coniferous forests versus only 6 insects and 2 diseases in deciduous forests. The operational efficiency demonstrated here enables continental-scale monitoring with near real-time disturbance alerts.

@@

\bibliographystyle{Frontiers-Harvard}
\bibliography{phenology}
